---
title: "Final Project Causal"
author: "Nidhal Jegham"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(fixest)
library(did)
library(bacondecomp)
library(dplyr)
library(ggplot2)
library(tidyr)
library(modelsummary)
library(patchwork)
```



## Preparing data

```{r}
df <- read.csv("C:/Users/jegha/OneDrive/Desktop/URI M1 S2/STA592 Causal/simulated_panel_data.csv")

df$cohort     <- ifelse(is.infinite(df$adoption_time), 0, df$adoption_time)
df$event_time <- ifelse(is.infinite(df$adoption_time), NA,
                        df$time - df$adoption_time)

print(paste0("Units:   ", n_distinct(df$id)))
print(paste0("Periods: ", n_distinct(df$time)))
print(table(df$cohort))
print(paste0("True ATT (all treated):  ", round(mean(df$Treatment_effect[df$D == 1]), 4)))
print(paste0("True ATT cohort g=40:    ", round(mean(df$Treatment_effect[df$D == 1 & df$cohort == 40]), 4)))
print(paste0("True ATT cohort g=60:    ", round(mean(df$Treatment_effect[df$D == 1 & df$cohort == 60]), 4)))
```

# Outcome Trends by Group

```{r}
trend_data <- df %>%
  group_by(time, group) %>%
  summarise(Y_mean = mean(Y), .groups = "drop")

p_trends<-ggplot(trend_data, aes(x = time, y = Y_mean,
                       color = group, linetype = group)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 40, linetype = "dashed", color = "gray40") +
  geom_vline(xintercept = 60, linetype = "dotted", color = "gray40") +
  annotate("text", x = 41, y = max(trend_data$Y_mean) * 0.97,
           label = "g=40", hjust = 0, size = 3.2, color = "gray30") +
  annotate("text", x = 61, y = max(trend_data$Y_mean) * 0.97,
           label = "g=60", hjust = 0, size = 3.2, color = "gray30") +
  scale_color_manual(
    values = c("control" = "black", "treated_1" = "#E41A1C", "treated_2" = "#377EB8"),
    labels = c("Never-treated", "Cohort g=40", "Cohort g=60")) +
  scale_linetype_manual(
    values = c("control" = "solid", "treated_1" = "solid", "treated_2" = "solid"),
    labels = c("Never-treated", "Cohort g=40", "Cohort g=60")) +
  labs(title = "Mean Outcome Over Time by Group",
       x = "Time", y = "Mean Y", color = "", linetype = "") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())

p_trends

ggsave("fig_trends.png", p_trends, width = 7, height = 4)

```

## TWFE Method

```{r}
twfe <- feols(Y ~ D | id + time,
              data = df,
              vcov = ~id)

summary(twfe)

twfe_coef <- as.numeric(coef(twfe)["D"])
twfe_se   <- as.numeric(se(twfe)["D"])
true_att  <- mean(df$Treatment_effect[df$D == 1])

print(paste0("TWFE estimate: ", round(twfe_coef, 4)))
print(paste0("True ATT:      ", round(true_att,  4)))
print(paste0("Bias:          ", round(twfe_coef - true_att, 4)))
```

```{r}
df_es <- df %>% filter(!is.na(event_time) | group == "control")

twfe_es <- feols(Y ~ i(event_time, ref = -1) | id + time,
                 data = df_es,
                 vcov = ~id)

twfe_es_coef <- coef(twfe_es)
twfe_es_ci   <- confint(twfe_es)

twfe_es_df <- data.frame(
  event_time = as.numeric(gsub("event_time::", "", names(twfe_es_coef))),
  estimate   = as.numeric(twfe_es_coef),
  conf.low   = as.numeric(twfe_es_ci[, 1]),
  conf.high  = as.numeric(twfe_es_ci[, 2])
)

p_twfe_es <- ggplot(twfe_es_df, aes(x = event_time, y = estimate)) +
  geom_point(size = 0.8) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "gray50") +
  labs(title = "TWFE Event Study",
       x = "Event Time", y = "Estimate") +
  theme_minimal(base_size = 12)
p_twfe_es

```

The TWFE estimate is 2.279 (SE $= 0.127$, $p < 0.001$), compared to the true ATT of 2.768, yielding a downward bias of $-0.489$ (approximately 18\%). Despite being highly statistically significant, the estimate is meaningfully incorrect. This bias arises because TWFE pools all $2\times2$ DiD comparisons, including those that use already-treated units as controls, as formalized by the Goodman-Bacon decomposition below. The TWFE event study additionally exhibits a pronounced downward slope in the pre-treatment period, with coefficients as large as $\sim 8$ at event time $-60$. This is not a genuine pre-trend violation. Under staggered adoption with heterogeneous treatment effects, the TWFE pre-period coefficients are contaminated by treatment effect heterogeneity itself: already-treated units used as implicit controls have experienced growing treatment effects, which mechanically distorts the pre-period estimates.


## Goodman-Bacon Decomposition Method

```{r}
df_bacon <- df %>%
  filter(cohort != 0) %>%                  
  bind_rows(df %>% filter(cohort == 0)) %>%
  select(id, time, Y, D)

bacon_out <- bacon(Y ~ D,
                   data      = df_bacon,
                   id_var    = "id",
                   time_var  = "time")

print(paste0("Bacon weighted avg: ", round(sum(bacon_out$estimate * bacon_out$weight), 4)))
print(paste0("TWFE estimate:      ", round(twfe_coef, 4)))

ggplot(bacon_out, aes(x = weight, y = estimate, shape = type, color = type)) +
  geom_point(size = 3) +
  geom_hline(yintercept = true_att, linetype = "dashed", color = "gray40") +
  annotate("text", x = max(bacon_out$weight) * 0.5,
           y = true_att + 0.1,
           label = "True ATT", size = 3.2, color = "gray40") +
  labs(title = "Goodman-Bacon Decomposition",
       x = "Weight", y = "2x2 DiD Estimate",
       shape = "Comparison", color = "Comparison") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```
The Goodman-Bacon decomposition breaks the TWFE estimate of 2.279 into three
$2\times2$ comparisons. The dominant comparison is treated vs.\ never-treated
(weight = 0.800, estimate = 2.537), which is clean and close to the true ATT.
The problematic comparison is later vs.\ earlier treated (weight = 0.103,
estimate = 0.034), where cohort $g=60$ uses the already-treated cohort $g=40$
as a control. tHIS comparison severely underestimates the true
effect. The earlier vs.\ later treated comparison (weight = 0.098, estimate =
2.529) is less distorted. The weighted average of these three components
exactly recovers the TWFE estimate of 2.279, confirming that the downward bias
is driven by the later vs.\ earlier treated comparison receiving non-trivial
weight with a near-zero estimate.


## Callaway-Sant'Anna Method

```{r}
cs <- att_gt(
  yname         = "Y",
  tname         = "time",
  idname        = "id",
  gname         = "cohort",
  xformla       = ~ x1 + x2 + x3,
  control_group = "nevertreated",
  data          = df,
  est_method    = "reg"
)


```

```{r}
cs_overall <- aggte(cs, type = "simple")
summary(cs_overall)

cs_cohort <- aggte(cs, type = "group")
summary(cs_cohort)

cs_es <- aggte(cs, type = "dynamic", na.rm = TRUE)
```

```{r}
p_cs_es <- ggdid(cs_es,
                 title = "Callaway-Sant'Anna Event Study",
                 xlab  = "Event Time",
                 ylab  = "ATT")

ggsave("fig_cs_es.png", p_cs_es, width = 7, height = 4)
p_cs_es
```

The Callaway-Sant'Anna estimator, using never-treated units as the control group and conditioning on covariates $x_1$, $x_2$, $x_3$, produces three levels of aggregation. The simple aggregation yields an overall ATT of 2.341 (SE $= 0.218$, 95\% CI: [1.915, 2.768]), while the dynamic aggregation yields 2.638 (SE $= 0.186$, 95\% CI: [2.272, 3.003]), substantially closer to the true value of 2.768. The difference arises because the dynamic aggregation upweights later event-time cells where treatment effects are larger, while the simple aggregation is dominated by the more numerous early post-treatment observations where effects are smaller. Both are valid estimands; the dynamic aggregation better reflects the average effect over the full post-treatment horizon. At the cohort level, the estimated ATT for cohort $g=40$ is 3.088 (SE $= 0.253$) against a true value of 3.556, and for cohort $g=60$ is 1.230 (SE $= 0.250$) against a true value of 1.597. Both estimates are statistically significant and directionally correct. The remaining bias ($-0.468$ for $g=40$, $-0.367$ for $g=60$) reflects both finite-sample variance, with 30 treated units per cohort, moderate estimation error is expected, and potential misspecification of the outcome regression model.

The event-study aggregation confirms two key features of the DGP. First, all pre-treatment coefficients (event times $-58$ through $-1$) are small and statistically indistinguishable from zero, supporting the parallel trends assumption. Second, post-treatment estimates increase monotonically with event time from 1.054 at event time 0 to 4.788 at event time 58, consistent with the linearly growing treatment effects embedded in the simulation. This dynamic pattern is entirely missed by the single TWFE coefficient, underscoring the value of the event-study approach.

## Sun-Abraham Method


```{r}
sa <- feols(Y ~ sunab(adoption_time, time) | id + time,
            data = df,
            vcov = ~id)


sa_agg <- aggregate(sa, agg = "ATT")
sa_agg

```

```{r}
sa_coef <- coef(sa)
sa_ci   <- confint(sa)

sa_es_df <- data.frame(
  event_time = as.numeric(gsub("time::", "", names(sa_coef))),
  estimate   = as.numeric(sa_coef),
  conf.low   = as.numeric(sa_ci[, 1]),
  conf.high  = as.numeric(sa_ci[, 2])
)

p_sa <- ggplot(sa_es_df, aes(x = event_time, y = estimate)) +
  geom_point(size = 0.8) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "gray50") +
  labs(title = "Sun-Abraham Event Study",
       x = "Event Time", y = "Estimate") +
  theme_minimal(base_size = 12)

ggsave("fig_sa_es.png", p_sa, width = 7, height = 4)
```





The Sun-Abraham estimator yields an overall ATT of 2.315 (SE $= 0.189$, $p < 0.001$), against a true value of 2.768, for a bias of $-0.453$. The interaction-weighted event study reveals the same dynamic structure recovered by Callaway-Sant'Anna: post-treatment estimates grow monotonically from 1.068 at event time 0 to 4.699 at event time 58. Several individual pre-period coefficients reach nominal significance (e.g., \texttt{time::-34}, $p = 0.0002$; \texttt{time::-11}, $p = 0.001$). However, with 59 pre-period coefficients estimated simultaneously, approximately three false rejections are expected at the 5\% level under the null by chance alone. The significant pre-period estimates are scattered across the event-time window with no systematic trend, and their magnitudes ($|estimate| < 0.81$) are small relative to the post-treatment estimates which range from 1.07 to 4.70. We interpret these as noise rather than evidence of pre-existing trends. The close agreement between the Sun-Abraham and Callaway-Sant'Anna event-study profiles cross-validates both estimators.


## Overlay Event-Study Comparison

```{r}
true_es <- df %>%
  filter(D == 1) %>%
  group_by(event_time) %>%
  summarise(estimate = mean(Treatment_effect), .groups = "drop") %>%
  mutate(method = "True ATT", conf.low = estimate, conf.high = estimate)

cs_es_df <- data.frame(
  event_time = cs_es$egt,
  estimate   = cs_es$att.egt,
  conf.low   = cs_es$att.egt - 1.96 * cs_es$se.egt,
  conf.high  = cs_es$att.egt + 1.96 * cs_es$se.egt,
  method     = "Callaway-Sant'Anna"
)

twfe_es_coef <- coef(twfe_es)
twfe_es_ci   <- confint(twfe_es)

twfe_es_df <- data.frame(
  event_time = as.numeric(gsub("event_time::", "", names(twfe_es_coef))),
  estimate   = as.numeric(twfe_es_coef),
  conf.low   = as.numeric(twfe_es_ci[, 1]),
  conf.high  = as.numeric(twfe_es_ci[, 2]),
  method     = "TWFE"
)
sa_coef <- coef(sa)
sa_ci   <- confint(sa)

sa_es_df <- data.frame(
  event_time = as.numeric(gsub("time::", "", names(sa_coef))),
  estimate   = as.numeric(sa_coef),
  conf.low   = as.numeric(sa_ci[, 1]),
  conf.high  = as.numeric(sa_ci[, 2]),
  method     = "Sun-Abraham"
)

combined_es <- bind_rows(true_es, cs_es_df, twfe_es_df, sa_es_df) %>%
  filter(event_time >= -10, event_time <= 40)

ggplot(combined_es, aes(x = event_time, y = estimate,
                         color = method, linetype = method)) +
  geom_line(linewidth = 0.8) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high, fill = method),
    alpha = 0.1, color = NA,
    data = combined_es %>% filter(method != "True ATT")
  ) +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = 0,    linetype = "dotted",  color = "gray50") +
  scale_color_manual(values = c(
    "True ATT"           = "black",
    "TWFE"               = "#E41A1C",
    "Callaway-Sant'Anna" = "#377EB8",
    "Sun-Abraham"        = "#4DAF4A")) +
  scale_linetype_manual(values = c(
    "True ATT"           = "solid",
    "TWFE"               = "longdash",
    "Callaway-Sant'Anna" = "solid",
    "Sun-Abraham"        = "twodash")) +
  scale_fill_manual(
    values = c("True ATT" = NA, "TWFE" = "#E41A1C",
               "Callaway-Sant'Anna" = "#377EB8", "Sun-Abraham" = "#4DAF4A"),
    guide = "none") +
  labs(title = "Event-Study Comparison",
       x = "Event Time", y = "Estimated ATT",
       color = "", linetype = "") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Summary Comparison Table



```{r}
true_overall <- mean(df$Treatment_effect[df$D == 1])
true_g40     <- mean(df$Treatment_effect[df$D == 1 & df$cohort == 40])
true_g60     <- mean(df$Treatment_effect[df$D == 1 & df$cohort == 60])

cs_overall_est <- cs_overall$overall.att
cs_overall_se_ <- cs_overall$overall.se
cs_g40_est     <- cs_cohort$att.egt[cs_cohort$egt == 40]
cs_g60_est     <- cs_cohort$att.egt[cs_cohort$egt == 60]
cs_g40_se_     <- cs_cohort$se.egt[cs_cohort$egt == 40]
cs_g60_se_     <- cs_cohort$se.egt[cs_cohort$egt == 60]

sa_overall_est <- sa_agg["ATT", "Estimate"]
sa_overall_se_ <- sa_agg["ATT", "Std. Error"]

results_table <- data.frame(
  Method       = c("True ATT", "TWFE", "Callaway-Sant'Anna", "Sun-Abraham"),
  Overall_Est  = round(c(true_overall, twfe_coef,       cs_overall_est, sa_overall_est), 3),
  Overall_SE   = round(c(NA,           twfe_se,         cs_overall_se_, sa_overall_se_), 3),
  Overall_Bias = round(c(0,
                          twfe_coef       - true_overall,
                          cs_overall_est  - true_overall,
                          sa_overall_est  - true_overall), 3),
  g40_Est      = round(c(true_g40, NA, cs_g40_est, NA), 3),
  g40_SE       = round(c(NA,       NA, cs_g40_se_, NA), 3),
  g40_Bias     = round(c(0,        NA, cs_g40_est - true_g40, NA), 3),
  g60_Est      = round(c(true_g60, NA, cs_g60_est, NA), 3),
  g60_SE       = round(c(NA,       NA, cs_g60_se_, NA), 3),
  g60_Bias     = round(c(0,        NA, cs_g60_est - true_g60, NA), 3)
)


results_table
```


This table summarizes the ATT estimates across all three methods against the known true values. At the aggregate level, all three estimators underestimate the true ATT of 2.768, with biases of $-0.489$ (TWFE), $-0.453$ (Sun-Abraham), and $-0.427$ (Callaway-Sant'Anna) under simple aggregation. However, the CS dynamic aggregation yields a bias of only $-0.130$, substantially closer to the truth, as it upweights later event-time cells where treatment effects are larger. The key distinction across methods emerges at the cohort level, where Callaway-Sant'Anna recovers estimates of 3.088 for cohort $g=40$ and 1.230 for cohort $g=60$, both statistically significant and directionally correct relative to the true values of 3.556 and 1.597 respectively. TWFE produces a single pooled coefficient and cannot decompose effects by cohort. Sun-Abraham, while recovering a similar overall ATT of 2.315 and correctly tracing the dynamic event-study profile, does not natively aggregate to cohort-specific ATTs in the same form as CS. Overall, the table illustrates that while all methods recover the broad direction of the treatment effect, only Callaway-Sant'Anna reveals the substantial heterogeneity across adoption cohorts that TWFE obscures.

## CS Cohort-Specific ATT Plot


```{r}
p_cs_cohort <- ggdid(cs, ylim = c(-1, 7),
                     title = "ATT(g,t): Callaway-Sant'Anna by Cohort") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
p_cs_cohort

ggsave("fig_cs_cohort.png", p_cs_cohort, width = 9, height = 5)
```


## Appendix

### Callaway-Sant'Anna Method Tables


```{r}
summary(cs)
```




```{r}
summary(cs_es)

```


## Sun-Abraham Method Table

```{r}
summary(sa)

```

